---
title: "Forecasting the 2024 U.S. Presidential Election"
subtitle: "Kamala Harris Projected to Defeat Donald Trump 48.3% to 47.1% in the Popular Vote and 270 to 268 in the Electoral College Based on Poll of Polls and Bayesian Modeling Analysis"
author: 
  - Talia Fabregas
  - Aliza Mithwani
  - Fatimah Yunusa
thanks: "The code and data used to perform this presidential election forecast can be found at: [https://github.com/taliafabs/USPresidentialPollingForecast2024.git](https://github.com/taliafabs/USPresidentialPollingForecast2024.git)."
date: today
date-format: long
abstract: "The U.S. Presidential election will take place on Tuesday, November 5th with Vice President Kamala Harris and former President Donald Trump in a close race for the White House. In this paper, we used the poll-of-polls method and applied a Bayesian model to estimate the winner of the popular vote and the winner of each of the seven battleground states: Arizona, Georgia, Nevada, North Carolina, Michigan, Wisconsin, and Pennsylvania, and Nebraska's second congressional disctict. Using the results of our poll-of-polls analysis, we predict that Vice President Harris will win the popular vote, 48.3% to 47.1% and the electoral college 270 to 268 by winning three of the seven battleground states, Wisconsin, Michigan, and Pennsylvania. Our analysis shows that the race is extremely tight and former President Trump winning the popular vote, electoral college, or both is well within the margin of error. Our results show a statistical tie when we account for error, bias, weaknesses, and limitations."
format: pdf
number-sections: true
bibliography: references.bib 
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false

# lines 24-26 were written with the help of ChatGPT
required_packages <- c("tidyverse", "janitor", "ggplot2", "scales", "rstanarm",  "bayesplot", "splines", "knitr", "kableExtra", "arrow", "here")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

library(tidyverse)
library(janitor)
library(ggplot2)
library(scales)
library(rstanarm)
library(bayesplot)
library(splines)
library(knitr)
library(kableExtra)
library(here)

# line 38 was written with the help of ChaatGPT
setwd(here::here())
analysis_data <- arrow::read_parquet("data/02-analysis_data/analysis_data.parquet")
harris_analysis_data <- arrow::read_parquet("data/02-analysis_data/harris-analysis_data.parquet")
trump_analysis_data <- arrow::read_parquet("data/02-analysis_data/trump-analysis_data.parquet")
```

# Introduction

<!-- Overview paragraph -->

On Tuesday November 5, 2024, Americans will head to the polls to elect their 47th president. Polling data has shown a tight race between VIce President Kamala Harris and former President Donald Trump since President Joe Biden made the historic and unprecedented decision to end his re-election campaign on July 21, 2024. President Biden immediately endorsed Vice President Harris on July 21, making her the presumptive Democratic nominee. Vice President Harris officially became the Democratic nominee on August 2, 2024 following a virtual rollcall. This paper uses presidential polling data from after President Joe Biden ended his re-election campaign and Vice President Harris became the presumptive Democratic nominee nominee and Bayesian analysis models to estimate the percentage of voters supporting Vice President Harris and former President Trump.

<!-- Estimands paragraph -->

The estimands, defined by @tswd as the values we are trying to estimate, are the percentages of voters that support Vice President Harris and former President Trump at the national level, in each of the seven battleground states (Arizona, Nevada, Georgia, North Carolina, Wisconsin, Michigan, and Pennsylvania), and in Nebraska's second congressional district.

<!-- Results paragraph -->

Based on the results of applying our Bayesian model, we estimate that Vice President Harris will receive 48.3% of the popular vote and win the swing states of Michigan, Wisconsin, Pennsylvania, and Nebraska CD-2, and former President Trump will receive 47.1% of the popular vote and win the swing states of Arizona, Nevada, North Carolina, and Georgia. We use a baseline of 222 electoral votes for Vice President Harris from safe and likely Democratic states that were won by both Hillary Clinton and President Biden in 2016 and 2020, and 219 electoral votes for former President Trump from safe and likely Republican states that he won in 2016 and 2020 [@270towin]. U.S. elections are decided by the electoral college, not the popular vote; the candidate who wins the popular vote is not guaranteed to win the election [@270towin]. We estimate that Vice President Harris will defeat former President Trump in the electoral college 270 to 268.

<!-- Why it matters paragraph -->

Political polarization in the United States is at an all-time high. Vice President Harris and former President Trump have presented staunchly different policies and visions for the future of the United States. Vice President Harris has campaigned on reproductive rights, supporting small businesses, building more affordable housing, expanding medicare coverage, and cutting taxes for the middle class. Former President Trump has campaigned on tariffs, sales taxes ending foreign aid to Ukraine, and securing the Southern border despite instructing his allies in the U.S. Senate to oppose a bipartisan border bill. He is has not yet acknowledged that he lost the 2020 election to President Joe Biden. Women make up 50% of the U.S. population, but all 46 presidents so far have been men. Only one person of color, Barack Obama, has ever been elected president.

The remainder of this paper is structured as follows. @sec-data contains an overview of the polling dataset from @fivethirtyeight that was used, visualizations of different variables, and summary statistics. @sec-model contains the Bayesian regression models used to predict the percentages of voters that will support Vice President Harris and former President Trump. @sec-results contains tables and visualizations that present the national popular vote and state-level results after applying the Bayesian regression model model. @sec-discussion contains detailed discussions about the .... [Appendix -@sec-idealized-meth] contains a thorough discussion about an idealized methodology that we would use if given a \$100,000 budget to build a survey to forecast the 2024 U.S. election and our idealized survey questions. [Appendix -@sec-pollster-meth] contains a deep-dive into and evaluation of the methodologies used in the Siena College/New York Times poll. The statistical programming language [@citeR] and the `tidyverse`, `janitor`, `ggplot`, `kableExtra`, `arrow`, `rstanarm`, and `spline` packages were used to perform this election forecast, clean the dataset, build the Bayesian regression models, create data visualizations and tables, and apply the model.

# Data {#sec-data}

## Overview

The data was downloaded on October 29, 2024; polling data released after this date was not considered anywhere in this paper. The presidential polls dataset from FiveThirtyEight contains national polls and state-level polls for each of the 50 states and congressional districts in Maine and Nebraska [@fivethirtyeight]. The polls are conducted by various pollsters, including YouGov, Siena/NYT, CES/YouGov, Marquette Law School, The Washington Post, and McCourtney Institute/YouGov. We analyze national data and state-level data for the seven swing states that are expected to determine the results of this election: Michigan, Wisconsin, Pennsylvania, North Carolina, Georgia, Nevada, and Arizona.

We cleaned this dataset to only include high-quality polls at the national and state-level conducted on or after July 21, 2024, where the `population` is likely voters. We determined which polls were high-quality based on the numeric grade, and cleaned the dataset to include only polls with a numeric grade of at least 3.0. Polls with a rating (`numeric_grade`) of 3.0 are conducted by the best pollsters in the United States and score in the 99th percentile or better for accuracy and transparency [@MorrisBestPollsters]. If a poll in the dataset from @fivethirtyeight includes multiple populations, the narrowest one is used (likely voters over registered voters and registered voters over all American adults) [@Morris538]. Additional data cleaning details are available in [Appendix -@sec-data-cleaning].

## Measurement

<!-- Some paragraphs about how we go from a phenomena in the world to an entry in the dataset. -->

During U.S. presidential and midterm election cycles, different pollsters conduct polls (surveys) to estimate support for Democratic, Republican, and third-party candidates and forecast the election results. Sampling and survey methodologies vary by pollster and we discuss the methodologies used by the Siena College/New York Times poll in [Appendix -@sec-pollster-meth]. Each poll aims to survey a population that is representative of the U.S. electorate or likely voters to predict the percentage of voters who will support each candidate and which candidate will win.

Our dataset from @fivethirtyeight is a a collection of presidential polls from different pollsters that were conducted during the 2024 presidential election cycle. Each entry represents the percentage of respondents to a unique poll supporting Vice President Harris (after July 21, 2024) or President Joe Biden (before July 21, 2024), former President Trump, and third-party candidates. Unique polls are identified by a poll id and each entry contains information about the poll, such as the pollster that conducted it, its population, sample size, and the methodology that was used and information about its quality and accuracy, including its numeric grade, pollscore, and transparency score.

## Outcome and predictor variables

We will use `end_date` (the date that a poll was completed), `state`, `pollster`, and `pollscore` to predict support for Vice President Harris and former President Trump at the national level, at the state level for each of the seven battleground states, and in Nebraska's second congressional district in @sec-model. The tables and visualizations below present possible relationships between the predictor variables and support for Vice President Harris and former President Trump.

### Variation in support for Harris and Trump by state

The percentage of voters that support Vice President Harris and former President Trump varies by state; in some states support for Vice President Harris is higher than the national average and in others, it is lower than the national average. The 2024 presidential election is expected to be decided by seven swing states: Arizona, Georgia, Nevada, North Carolina, Wisconsin, Michigan, and Pennsylvania and Nebraska's second congressional district [@270towin]. Two states, Maine and Nebraska, award one electoral vote to the popular vote winner in each congressional district and an additional two electoral votes to the statewide popular vote winner [@270towin].

```{r}
#| label: tbl-avgsummary
#| tbl-cap: Polling averages for Harris and Trump at the national level and at the state level for the states included in the polling dataset show a narrow popular vote lead for Vice President Harris and extremely close races in the 7 battleground states (Arizona, Georgia, Nevada, North Carolina, Michigan, Pennsylvania, Wisconsin as of October 29, 2024.
#| echo: false
#| warning: false
#| message: false

analysis_data_harris <- analysis_data |>
  filter(end_date >= as.Date("2024-07-21"))

swing_states <- c("Arizona", "Georgia", "Nevada", "Pennsylvania", "Michigan", "Wisconsin", "North Carolina")

polling_averages <- analysis_data_harris %>%
  filter(state == "National" | state %in% analysis_data_harris$state) %>%
  group_by(state) %>%
  filter(end_date == max(end_date)) %>% 
  mutate(
    pct_harris = if_else(candidate_name == "Kamala Harris", pct, NA_real_),
    pct_trump = if_else(candidate_name == "Donald Trump", pct, NA_real_)
  ) %>%
  summarise(
    Harris = mean(pct_harris, na.rm = TRUE),
    Trump = mean(pct_trump, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # Rename National to Popular Vote
  mutate(state = if_else(state == "National", "Popular Vote", state)) %>%
  # Reorder to ensure Popular Vote is first
  arrange(match(state, c("Popular Vote", swing_states)))

# Create the kable with the modified state names
polling_table <- polling_averages %>%
  select(State = state, `Harris %` = Harris, `Trump %` = Trump)

# Display the table
kable(polling_table, format = "html", digits = 2) 
```

Based on our presidential polling data from @fivethirtyeight Vice President Harris leads former President Trump in the popular vote 50.5% to 48.0%. The margins in the seven battleground states are tight, with Vice President Harris leading in Nevada, Michigan, Wisconsin, and Nebraska's second congressional district and former President Trump leading in Arizona, Georgia, North Carolina. The two candidates are tied in Pennsylvania. The data set also includes state-level polls from likely Democratic states (Minnesota, New Hampshire, Virginia) and likely Republican states (Florida, Missouri, Montana, Nebraska, Ohio, Texas) that are not expected to determine the winner of the election [@270towin]. Support percentages for Harris and Trump are closer, but still have some variation among the seven battleground states. Harris' support in five of the seven battleground states is lower than her national support, while Trump's support in four of the seven battleground states is higher than his national support.

Since President Biden ended his re-election campaign and Vice President Harris became the Democratic Presidential nominee, the polls have shown a close race between Vice President Harris and former President Trump. Polling averages for the six months leading up to election day, including from before President Biden withdrew on July 21, 2024 can be found in [Appendix -@sec-Biden]. @fig-nationalavg2 shows national polling averages for Harris and Trump since July 21 and @fig-sevenbattlegroundskamala shows state-level polling averages fo Harris and Trump in the seven battleground states and Nebraska's second congresional district.

Vice President Harris surpassed former President Trump in popular vote polls shortly after becoming the presumptive Democratic nominee in late July, but her lead narrowed in mid-August and the polls have been neck-and-neck since late August.

```{r}
#| label: fig-nationalavg2
#| fig-cap: Harris leads Trump in national popular vote polling. The lines are moving poll averages and the points are individual high-quality polls. Color shading of points indicates which candidate won the poll.
#| echo: false
#| warning: false
#| message: false

national2 <- analysis_data |>
  filter(state=="National",
         (candidate_name == "Kamala Harris" | candidate_name == "Donald Trump"),
         end_date >= as.Date("2024-07-21")
         )

ggplot(national2, aes(x = end_date, y = pct, color = candidate_name)) +
  stat_summary(geom = "line", fun = mean) +      # Plot the mean line for polling percentages
  geom_point(size = 1, alpha = 0.25) +            # Add points for each individual poll
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  # geom_vline(xintercept = as.Date("2024-07-21"), linetype = "dashed", color = "darkgray") +
  labs(title = "",
       x = "",
       y = "",
       color = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(limits = c(0, 60), labels = scales::percent_format(scale = 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %e")
```

```{r}
#| label: fig-sevenbattlegroundskamala
#| fig-cap: Harris and Trump have been in a dead heat in swing state polls since July 2024. The lines are moving poll averages for each candidate in the seven battleground states. The points are individual state-level polls. The color of each point indicates which candidate lead in the poll.
#| echo: false
#| warning: false
#| message: false

sevenbattlegrounds2 <- analysis_data |>
  filter((state=="Wisconsin" | state == "Michigan" | state == "Pennsylvania" |
           state == "Nevada" | state == "Arizona" | state == "Georgia" | 
           state == "North Carolina" | state == "Nebraska CD-2"),
         (candidate_name == "Kamala Harris" | candidate_name == "Donald Trump"),
         end_date >= as.Date("2024-07-21"))

ggplot(sevenbattlegrounds2, aes(x = end_date, y = pct, color = candidate_name)) +
  stat_summary(geom = "line", fun = mean) +      # Plot the mean line for polling percentages
  geom_point(size = 1, alpha = 0.15) +            # Add points for each individual poll
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  labs(title = "",
       x = "",
       y = "",
       color = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(limits = c(0, 60), labels = scales::percent_format(scale = 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %e") +
  facet_wrap(~state)
```

### Variation in support for Harris and Trump by pollster and pollscore

Talk more about it.

```{r}
#| label: fig-pollsters2
#| fig-cap:  Harris and Trump polling averages vary by pollster. The lines are moving poll averages for each candidate in the seven battleground states. The points are individual polls. The color of each point indicates which candidate lead in the poll.
#| echo: false
#| warning: false
#| message: false

harris_versus_trump <- analysis_data |>
  filter(end_date >= as.Date("2024-07-21"))

ggplot(national2, aes(x = end_date, y = pct, color = candidate_name)) +
  stat_summary(geom = "line", fun = mean) +      
  geom_point(size = 1, alpha = 0.25) +           
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  labs(title = "",
       x = "",
       y = "",
       color = "") +
  theme_minimal() +
  theme(strip.text = element_text(size = 8)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(limits = c(0, 60), labels = scales::percent_format(scale = 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %e") +
  facet_wrap(~pollster)
```

@tbl-pollscore-table shows the different pollscores that appear in our dataset and how often they appear.

```{r}
#| label: tbl-pollscore-table
#| tbl-cap: Only high-quality datasets with a numeric grade of at least 3.0 were included, so the polls included all have good pollscores. Negative/lower pollscores are better.
#| echo: false
#| warning: false
#| message: false

# subset the analysis data to only include harris and trump after biden dropped out
analysis_data2 <- analysis_data |>
  filter((candidate_name == "Kamala Harris" | candidate_name == "Donald Trump"),end_date >= as.Date("2024-07-21")) 

pollscore_counts <- analysis_data2 %>%
  group_by(pollscore) %>%
  summarize(num_polls = n())

# Display the table
kable(pollscore_counts, col.names = c("Pollscore", "Number of Polls"))
```

```{r}
#| label: tbl-pollscore-pollsters
#| tbl-cap: All the pollsters included in our analysis dataset have a numeric grade of 3, but Siena/NYT has the best pollscore at -1.5.
#| echo: false
#| warning: false
#| message: false

avg_pollscore_table <- analysis_data2 %>%
  group_by(pollster) %>%
  summarize(
    avg_pollscore = mean(pollscore, na.rm = TRUE),
    numeric_grade = mean(numeric_grade, na.rm=TRUE)) %>%
  arrange(desc(avg_pollscore))

# Display the table
kable(avg_pollscore_table, col.names = c("Pollster", "Pollscore", "Numeric Grade"))
```

# Model {#sec-model}

The goal of our modeling strategy is to use a Bayesian model to investigate the relationship between the percentage of voters in a poll who support Vice President Harris or Former President Trump and the date the poll was conducted, the state (or if it was a national poll), the pollster who conducted the poll, and the pollscore it received. Model details and diagnostics are presented in [Appendix -@sec-model-details]. We use two Bayesian regression models, one each to model the percentage of voters supporting Vice President Harris, $pct\_harris$ and the percentage of voters supporting former President Trump, $pct\_trump$.

The models to predict $pct\_harris$ and $pct\_trump$ both use the following predictors:

-   `end_date_num`: the number of days since July 21, 2024, when President Biden ended his re-election campaign and endorsed Vice President Harris. This is the spline term; it uses an spline with 5 degrees of freedom to model changes in $pct\_harris$ (in the Harris model) and $pct\_trump$ (in the Trump model) over time.

-   `state`: accounts for the change in $pct\_harris$ and $pct\_trump$ at the state-level for a particular state or at the national level.

-   `pollster`: accounts for the differences in $pct\_harris$ and $pct\_trump$ among different pollsters.

-   `pollscore:`adjusts for the pollscore rating, which is the Predictive Optimization of Latent skill Level in Surveys, Considering Overall Record, Empirically, calculated by averaging the predictive error and predictive bias of a poll [@Morris538].

## Model set-up {#sec-modelsetup}

Define $pct\_harris_i$ as the percentage of voters supporting Vice President Harris and $pct\_trump_i$ as the percentage of voters supporting former President Trump in the poll with unique `poll_id`=*i*.

Harris model:

```{=tex}
\begin{align*} 
pct\_harris_i &= \beta_0 + \beta_1 \cdot ns(end\_date\_num_i, df=5) + \beta_2 \cdot state_i + \beta_3 \cdot pollster_i + \beta_4 pollscore_i \\
\beta_0 &\sim \mbox{Normal}(50, 10) \\
\beta_1 &\sim \mbox{Normal}(0, 5) \\
\beta_2 &\sim \mbox{Normal}(0, 5) \\
\beta_3 &\sim \mbox{Normal}(0, 5) \\
\beta_4 &\sim \mbox{Normal}(0, 5) 
\end{align*}
```
Trump model:

```{=tex}
\begin{align*} 
pct\_trump_i &= \beta_0 + \beta_1 \cdot ns(end\_date\_num_i, df=5) + \beta_2 \cdot state_i + \beta_3 \cdot pollster_i + \beta_4 pollscore_i \\
\beta_0 &\sim \mbox{Normal}(50, 10) \\
\beta_1 &\sim \mbox{Normal}(0, 5) \\
\beta_2 &\sim \mbox{Normal}(0, 5) \\
\beta_3 &\sim \mbox{Normal}(0, 5) \\
\beta_4 &\sim \mbox{Normal}(0, 5) 
\end{align*}
```
The model is run in R [@citeR] using the `rstanarm` package of @rstanarm. The default priors from `rstanarm` are used for both GLM Bayesian models. The intercept normal prior with $\mu=50$ and $\sigma=10$ reflects the central tendencies of Harris and Trump's polling percentages, influenced by prior knowledge and the predictors use a normal prior with $\mu=0$ and $\sigma=5$ [@rstanarm].

### Model justification

We use separate Bayesian regression models with the same predictors to estimate $pct\_harris$ and $pct\_trump$.

because using only one model to predict support for Harris ($pct\_harris$) and calculating support for Trump, $pct\_trump$, as 100% - $pct\_harris$, or vice versa, would be insufficient and lead to inaccurate results because there are third-party candidates who will receive a small percentage of the vote nationally and in each of the battleground states. It is possible for either candidate to win the popular vote or a swing state with less than 50% of the vote because of third-party candidates, and simply assuming that $pct\_trump$ = 1 - $pct\_harris$ (or vice versa) would not account for this. In 2020, President Biden won Arizona with and Georgia with. In 2016, Hillary Clinton beat former President Trump in the popular vote % to %.

We used the default priors from `rstanarm` in both our models because using a normal prior with $\mu=50$ and $\sigma=10$ for the intercept of each model allows us to reflect the central tendencies of $pct\_harris$ and $pct\_trump$ and the prior knowledge from the @fivethirtyeight presidential polls that show a close race between Vice President Harris at the national popular vote level and in each of the seven battleground states [@rstanarm].

Model validation and checking results can be found in [Appendix -@sec-model-details]

### Alternative model considered

We initially used only one Bayesian regression model to predict $pct\_harris$ with `end_days_num` (spline term), `state`, `pollster`, and `pollscore` as predictors. We then calculated $pct\_trump$ as 1 - $pct\_harris$. This approach was insufficient and may have produced misleading or inaccurate results because there are third-party candidates who will receive a small percentage of the vote nationally and in each state, so Harris or Trump could win a state or the popular vote with less than 50%.

### Model weaknesses and limitations

A limitation of our model is ...Support for Harris at a given time is not a predictor for support for Trump and vice versa The models used to predict support for Harris and Trump do not consider support for the other candidate at any given time. Support for Harris likely has a negative relationship with support for Trump; when support for Harris increases, support for Trump likely decreases. Third party candidates also matter when quantifying this and our models do not consider them at all.

Another weakness in our model is the fact that it does not give higher-quality polls more weight or influence. Our A next step might be to give polls with a larger sample size or better pollscore more influence.

# Results {#sec-results}

<!-- The model results and model summary are presented in [Appendix -@sec-model-details]. -->

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

harris_model <-
  readRDS(file = here::here("models/harris_model.rds"))

trump_model <- 
  readRDS(file = here::here("models/trump_model.rds"))
```

## National Popular Vote Results

<!-- Popular vote prediction after applying the Harris and Trump models. -->

We applied the Harris model and the Trump model defined in @sec-modelsetup to predict each candidate's share of the national popular vote.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(splines)

harris_analysis_data$state <- factor(harris_analysis_data$state)
harris_analysis_data$pollster <- factor(harris_analysis_data$pollster)
harris_analysis_data$pollscore <- as.numeric(as.character(harris_analysis_data$pollscore))

# Chat GPT was used to help write this code
# Define the unique levels of pollster
pollster_levels <- levels(harris_analysis_data$pollster)
pollscore_levels <- unique(harris_analysis_data$pollscore)

# Initialize a list to store predictions for each pollster
all_pollster_preds <- list()

# Loop over each pollster to generate predictions
for (p in pollster_levels) {
  # Create new data for each pollster
  random_pollscore <- sample(c(-1.1, -1.5, -1.2), 1)
  harris_new_data <- data.frame(
    end_date_num = seq(
      min(harris_analysis_data$end_date_num),
      max(harris_analysis_data$end_date_num),
      length.out = 100
    ),
    pollster = factor(p, levels = pollster_levels),
    state = "National",
    # state = factor("National", levels = levels(harris_analysis_data$state)),
    pollscore = random_pollscore
  )
  
  # Generate posterior predictions for this pollster
  preds <- posterior_predict(harris_model, newdata = harris_new_data)
  all_pollster_preds[[p]] <- preds  # Store in list
}

# Combine predictions across all pollsters
# Convert list of matrices into a 3D array: [iterations, predictions, pollster]
all_pollster_preds_array <- simplify2array(all_pollster_preds)

# Calculate summary statistics across pollsters
# Averaging predictions across pollsters at each iteration
pred_mean <- apply(all_pollster_preds_array, c(1, 2), mean)
pred_lower <- apply(all_pollster_preds_array, 2, quantile, probs = 0.025)
pred_upper <- apply(all_pollster_preds_array, 2, quantile, probs = 0.975)

# Combine these into a summary data frame
harris_pred_summary <- harris_new_data %>%
  mutate(
    pred_mean = colMeans(pred_mean),
    pred_lower = pred_lower,
    pred_upper = pred_upper
  )

harris_spline_plot <- ggplot(harris_analysis_data, aes(x = end_date_num, y = pct_harris, color = pollster)) +
    geom_point() +
    geom_line(
      data = harris_pred_summary,
      aes(x = end_date_num, y = pred_mean),
      color = "blue",
      inherit.aes = FALSE
    ) +
    geom_ribbon(
      data = harris_pred_summary,
      aes(x = end_date_num, ymin = pred_lower, ymax = pred_upper),
      alpha = 0.2,
      inherit.aes = FALSE
    ) +
    labs(
      x = "Days since Biden withdrew",
      y = "Percentage",
      title = "Poll Percentage over Time with Spline Fit"
    ) +
    theme_minimal() +
    scale_y_continuous(labels = scales::percent_format(scale = 1)) 
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(splines)
trump_analysis_data$state <- factor(trump_analysis_data$state)
trump_analysis_data$pollster <- factor(trump_analysis_data$pollster)
trump_analysis_data$pollscore <- as.numeric(as.character(trump_analysis_data$pollscore))


# Define unique pollster levels
pollster_levels <- levels(trump_analysis_data$pollster)

# Initialize list to store predictions for each pollster
all_pollster_preds <- list()

# Loop over each pollster to generate predictions
for (p in pollster_levels) {
  # Create trump_new_data for each pollster
  random_pollscore <- sample(c(-1.1, -1.5, -1.2), 1)
  trump_new_data <- data.frame(
    end_date_num = seq(
      min(trump_analysis_data$end_date_num),
      max(trump_analysis_data$end_date_num),
      length.out = 100
    ),
    pollster = factor(p, levels = pollster_levels),
    state = "National",
    # state = factor("National", levels = levels(trump_analysis_data$state)),
    pollscore = random_pollscore
  )
  
  # Generate posterior predictions for this pollster
  preds <- posterior_predict(trump_model, newdata = trump_new_data)
  all_pollster_preds[[p]] <- preds  # Store in list
}

# Combine predictions across all pollsters
# Convert list of matrices into a 3D array: [iterations, predictions, pollster]
all_pollster_preds_array <- simplify2array(all_pollster_preds)

# Calculate summary statistics across pollsters
# Averaging predictions across pollsters at each iteration
pred_mean <- apply(all_pollster_preds_array, c(1, 2), mean)
pred_lower <- apply(all_pollster_preds_array, 2, quantile, probs = 0.025)
pred_upper <- apply(all_pollster_preds_array, 2, quantile, probs = 0.975)

# Combine these into a summary data frame
trump_pred_summary <- trump_new_data %>%
  mutate(
    pred_mean = colMeans(pred_mean),
    pred_lower = pred_lower,
    pred_upper = pred_upper
  )

# Plot using the aggregated predictions
trump_spline_plot <- ggplot(trump_analysis_data, aes(x = end_date_num, y = pct_trump, color = pollster)) +
  geom_point() +
  geom_line(
    data = trump_pred_summary,
    aes(x = end_date_num, y = pred_mean),
    color = "red",
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    data = trump_pred_summary,
    aes(x = end_date_num, ymin = pred_lower, ymax = pred_upper),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  labs(
    x = "Days since Biden withdrew",
    y = "",
    title = "Poll Percentage over Time with Spline Fit (Trump)"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) 
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| fig-label: fig-nationalresults
#| fig-cap: Results of applying Bayesian regression models with spline to predict support for Harris and Trump show that Harris is leading the popular vote 48.3% to 47.1% as of October 29. Lines are moving averages from applying the model, points are individual polls, and the shaded areas show the range of values within the margin of error.

# Add a candidate column with full names to each summary data frame
harris_pred_summary <- harris_pred_summary %>%
  mutate(candidate = "Kamala Harris")

trump_pred_summary <- trump_pred_summary %>%
  mutate(candidate = "Donald Trump")

# Combine the data frames
combined_pred_summary <- bind_rows(harris_pred_summary, trump_pred_summary)

# Plot the combined data with both candidates' lines and ribbons
ggplot() +
  # Plot the original data points for Harris
  geom_point(data = harris_analysis_data, aes(x = end_date_num, y = pct_harris, color = pollster), alpha = 0.6) +
  # Plot the original data points for Trump
  geom_point(data = trump_analysis_data, aes(x = end_date_num, y = pct_trump, color = pollster), alpha = 0.6) +
  # Plot the prediction line for Kamala Harris
  geom_line(
    data = combined_pred_summary %>% filter(candidate == "Kamala Harris"),
    aes(x = end_date_num, y = pred_mean, color = candidate),
    inherit.aes = FALSE
  ) +
  # Plot the prediction line for Donald Trump
  geom_line(
    data = combined_pred_summary %>% filter(candidate == "Donald Trump"),
    aes(x = end_date_num, y = pred_mean, color = candidate),
    inherit.aes = FALSE
  ) +
  # Plot the prediction ribbon for Kamala Harris
  geom_ribbon(
    data = combined_pred_summary %>% filter(candidate == "Kamala Harris"),
    aes(x = end_date_num, ymin = pred_lower, ymax = pred_upper, fill = candidate),
    alpha = 0.1,
    inherit.aes = FALSE
  ) +
  # Plot the prediction ribbon for Donald Trump
  geom_ribbon(
    data = combined_pred_summary %>% filter(candidate == "Donald Trump"),
    aes(x = end_date_num, ymin = pred_lower, ymax = pred_upper, fill = candidate),
    alpha = 0.1,
    inherit.aes = FALSE
  ) +
  annotate("text", x = max(combined_pred_summary$end_date_num) -0.5, y = 48.3, 
           label = "48.3", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_pred_summary$end_date_num) - 0.5, y = 47.1, 
           label = "47.1", color = "red", hjust = -0.2, size = 2) +  # Adjust size here
  labs(
    x = "Days since Biden withdrew",
    y = "",
    title = ""
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  scale_fill_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
      guides(
      color = guide_legend(
        title = "", 
        override.aes = list(fill = NA, linetype = 1)
      ),
      fill = "none"
    )
```

## State-Level Results for the Seven Battleground States and Nebraska's Second Congressional District #sec-42

Vice President Harris starts out with 225 electoral college votes from states that went to both Hillary Clinton in 2016 and Joe Biden in 2020 [@270towin]. These states are considered safe or likely Democrat and available poling from @fivethirtyeight shows Vice President Harris with a lead outside the five percent margin of error and in most cases, a double-digit lead. Former President Trump starts out with 219 electoral college votes from states that he won in both 2016 and 2020 [@270towin]. These states are considered safe or likely Republican and available polling data from @fivethirtyeight shows former President Trump with a lead outside the five percent margin of error, and in most cases, a double-digit lead. For this reason, the results of the Bayesian models presented in @sec-model focus on estimating support for Vice President Harris and former President Trump in Arizona, Nevada, Georgia, North Carolina, Wisconsin, Michigan, Pennsylvania, and Nebraska's second congressional district, which are worth a total of 94 electoral votes [@270towin].

We applied the Harris model and the Trump model defined in @sec-modelsetup at the state-level to predict the percentage of the vote that each candidate will receive in each of the seven battleground states. This will allow us to predict the winner of the electoral college and who will become the 47th President. @fig-swingresults shows the results of applying the Harris and Trump Bayesian models with spline at the state-level over time.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(dplyr)
library(ggplot2)
library(broom)

# Define the swing states
swing_states <- c("Nevada", "Arizona", "Georgia", "Michigan", "Pennsylvania", "North Carolina", "Wisconsin", "Nebraska CD-2")

# Initialize an empty list to store plots
all_plots <- list()
state_margins <- list()

# Loop over each swing state
for (state in swing_states) {
  # Filter data for the current state
  harris_state <- harris_analysis_data %>% filter(state == state | state == "National")
  trump_state <- trump_analysis_data %>% filter(state == state | state == "National")
  
  # Get pollster levels for the current state
  pollster_levels <- levels(harris_state$pollster)

  # Initialize a list to store predictions for Harris
  harris_preds <- list()

  # Generate predictions for Harris
  for (p in pollster_levels) {
    random_pollscore <- sample(c(-1.1, -1.5, -1.2), 1)
    harris_new_data <- data.frame(
      end_date_num = seq(min(harris_state$end_date_num), max(harris_state$end_date_num), length.out = 100),
      pollster = factor(p, levels = pollster_levels),
      state = state,
      # state = factor(state, levels = levels(harris_state$state)),
      pollscore = random_pollscore
    )
    
    preds <- posterior_predict(harris_model, newdata = harris_new_data)
    harris_preds[[p]] <- preds
  }

  # Combine predictions for Harris
  harris_preds_array <- simplify2array(harris_preds)
  harris_pred_mean <- apply(harris_preds_array, c(1, 2), mean)
  harris_pred_lower <- apply(harris_preds_array, 2, quantile, probs = 0.025)
  harris_pred_upper <- apply(harris_preds_array, 2, quantile, probs = 0.975)
  
  harris_pred_summary <- harris_new_data %>%
    mutate(pred_mean = colMeans(harris_pred_mean),
           pred_lower = harris_pred_lower,
           pred_upper = harris_pred_upper,
           candidate = "Kamala Harris")

  # Initialize a list to store predictions for Trump
  trump_preds <- list()
  
  # Generate predictions for Trump
  for (p in pollster_levels) {
    random_pollscore <- sample(c(-1.1, -1.5, -1.2), 1)
    trump_new_data <- data.frame(
      end_date_num = seq(min(trump_state$end_date_num), max(trump_state$end_date_num), length.out = 100),
      pollster = factor(p, levels = pollster_levels),
      state = state,
      # state = factor(state, levels = levels(trump_state$state)),
      pollscore = random_pollscore
    )
    
    preds <- posterior_predict(trump_model, newdata = trump_new_data)
    trump_preds[[p]] <- preds
  }

  # Combine predictions for Trump
  trump_preds_array <- simplify2array(trump_preds)
  trump_pred_mean <- apply(trump_preds_array, c(1, 2), mean)
  trump_pred_lower <- apply(trump_preds_array, 2, quantile, probs = 0.025)
  trump_pred_upper <- apply(trump_preds_array, 2, quantile, probs = 0.975)
  
  trump_pred_summary <- trump_new_data %>%
    mutate(pred_mean = colMeans(trump_pred_mean),
           pred_lower = trump_pred_lower,
           pred_upper = trump_pred_upper,
           candidate = "Donald Trump")

  margin <- mean(harris_pred_summary$pred_mean - trump_pred_summary$pred_mean)
  state_margins[[state]] <- margin
  # Combine predictions for both candidates
  combined_pred_summary2 <- bind_rows(harris_pred_summary, trump_pred_summary)

  # Plot for the current state
  plot <- ggplot() +
    geom_point(data = harris_state, aes(x = end_date_num, y = pct_harris, color = pollster), alpha = 0.6) +
    geom_point(data = trump_state, aes(x = end_date_num, y = pct_trump, color = pollster), alpha = 0.6) +
    geom_line(data = combined_pred_summary2 %>% filter(candidate == "Kamala Harris"),
              aes(x = end_date_num, y = pred_mean, color = candidate), inherit.aes = FALSE) +
    geom_line(data = combined_pred_summary2 %>% filter(candidate == "Donald Trump"),
              aes(x = end_date_num, y = pred_mean, color = candidate), inherit.aes = FALSE) +
    geom_ribbon(data = combined_pred_summary2 %>% filter(candidate == "Kamala Harris"),
                aes(x = end_date_num, ymin = pred_lower, ymax = pred_upper, fill = candidate),
                alpha = 0.1, inherit.aes = FALSE) +
    geom_ribbon(data = combined_pred_summary2 %>% filter(candidate == "Donald Trump"),
                aes(x = end_date_num, ymin = pred_lower, ymax = pred_upper, fill = candidate),
                alpha = 0.1, inherit.aes = FALSE) +
    labs(x = "Days since Biden withdrew", y = "", title = state) +
    theme_minimal() +
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
    scale_fill_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
    guides(color = guide_legend(title = "Candidate", override.aes = list(fill = NA, linetype = 1)),
           fill = "none")

  # Save the plot to the list
  all_plots[[state]] <- plot
}
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-swingresults
#| fig-cap: "The lines are the weighted moving averages resulting from applying the Bayesian model with spline, dots are individual state-level polls, and shaded areas show range of values within the margin of error."
#| fig-subcap: ["Avg Trump lead +1.61", "Avg Trump lead +3.01", "Avg Trump lead +0.37","Avg Trump lead +0.44", "Avg Harris lead +2.22", "Avg Harris lead +2.57", "Avg Harris lead +3.31", "Avg Harris lead +11.87"] 
#| layout-ncol: 2
#| layout-nrow: 4

az <- all_plots[["Arizona"]]
ga <- all_plots[["Georgia"]]
nv <- all_plots[["Nevada"]]
nc <- all_plots[["North Carolina"]]
mi <- all_plots[["Michigan"]]
pa <- all_plots[["Pennsylvania"]]
wi <- all_plots[["Wisconsin"]]
ne2 <- all_plots[["Nebraska CD-2"]]

print(az)
print(ga)
print(nv)
print(nc)
print(mi)
print(pa)
print(wi)
print(ne2)
```

@fig-swingresults shows former President Trump leading in Arizona, Nevada, Georgia, and North Carolina and Vice President Harris leading in Michigan, Wisconsin, Pennsylvania, and Nebraska's second congressional district as of October 29, 2024. The margins of error for Trump and Harris have little overlap for Nebraska CD-2, so we can predict that Harris will likely win its single electoral vote. Nebraska CD-2 will not be considered safe or likely Democratic because there is still overlap in the margins of error for Harris and Trump and it was not won by Clinton in 2016. The shaded margins of error for the seven battleground states have a lot of overlap; this means that it is within the margin of error for either candidate to win any of the seven battleground states.

\newpage

## Electoral College Results

Before the seven key battleground states are decided, Vice President Harris has 225 electoral votes from safe and likely Democratic states and former President Trump has 219 from safe and likely Republican states [@270towin]. We use this starting point and the results from @sec-42 to predict the electoral college. 

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| include: true
#| label: tbl-swingectable
#| tbl-cap: Electoral votes for the seven battleground states and Nebraska CD-2

library(tibble)
setwd("/Users/talia/USPresidentialPollingForecast2024")
electoral_votes <- read_csv("data/02-analysis_data/electoral_college_votes.csv")
electoral_votes <- electoral_votes %>%
  mutate(state = str_to_title(state))

swing_states <- c("Arizona", "Nevada", "Georgia", "North Carolina", "Michigan", "Wisconsin",
                  "Pennsylvania", "Nebraska-2")

swing_votes <- electoral_votes |>
  filter(state %in% swing_states)

swing_votes_tibble <- tibble(
  "State" = swing_votes$state,
  "Electoral Votes" = swing_votes$electoral_votes
)

kable(swing_votes_tibble)
```

If the results presented in @fig-swingresults hold on election night, and Harris and Trump win their safe and likely states, Harris would defeat Trump in the electoral college, 270 to 268. This is the mean estimate for the number of electoral college votes for Harris and Trump. If Vice President Harris wins all seven battleground states and Nebraska CD-2, she would defeat former President Trump in the electoral college 319 to 219. If former President Trump wins all seven battleground states, he would defeat Vice President Harris in the electoral college 313 to 225.

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| include: true
#| label: tbl-ec_results
#| tbl-cap: The electoral college outcomes that fall within the margin of error for the seven battleground states.

library(tibble)

ec_results <- tibble(
  " " = c("Harris Upper/Trump Lower Estimate", "Mean Estimate", "Harris Lower/Trump lower Estimate"),
  "Harris" = c(319, 270, 225),
  "Trump" = c(219, 268, 313)
)

kable(ec_results)
```

# Discussion {#sec-discussion}

## The margin of error and why Harris or Trump could win {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Polling flaws in 2016, 2020, and 2022

Discuss polling flaws in 2016 and 2020, how support for Trump was underestimated. Discuss how this was overcorrected in 2022, and how support for GOP house and senate candidates was overestimated. Discuss Trump's unique appeal to a certain segment of the electorate. Refer to Nate Silver articles.

## Something else about polling, nonresponse, and how a repeat of any of these flaws could swing the race either way

## Weaknesses, limitations, and next steps

Biden's withdrawal, use of two models, no third-party candidates, only national popular vote and swing states were analyzed, we did not consider how the pollsters in our dataset handled nonresponse when modeling. Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Idealized methodology {#sec-idealized-meth}

## Survey objectives

## Sampling approach

The target population for our idealized survey is likely voters in the United States. We will use the L2 voter database from @L2 to build a representative sampling frame. The L2 voter database is one of the most trusted sources for enhanced voter data and it includes detailed demographics and voting history data [@L2]. Using this data set is the first step towards ensuring that our sample aligns with the general electorate and is representative of likely voters so that we can sample precisely.

We will use a stratified sampling approach to closely examine voter demographics. Stratified sampling allows us to look at every stratum and carry out simple random sampling with those strata [@tswd]. Its main goal is to ensure that every strata of the population is represented [@Neyman1934]. We selected stratified sampling because it will allow us to have representation within the subgroups that we are interested in and it has a reduced sampling error and improved accuracy rate [@tswd]. We considered simple random sampling, but ultimately chose stratified sampling instead because it tends to produce more precise accurates when used to forecast U.S. elections [@pewhacking]. The U.S. election is decided by the electoral college, not the popular vote so we will oversample from the seven battleground states that are expected to decide the 2024 election: Arizona, Nevada, Georgia, North Carolina, Wisconsin, Michigan, and Pennsylvania [@fivethirtyeight]. Oversampling from the swing states will allow us to put an emphasis on forecasting their results and increase the accuracy of our electoral college estimate. Our target sample size of 100,000 respondents will provide a margin of error of 2% at a 95% confidence level; this will allow us to be precise while still accounting for budgetary constraints [@pewhacking].

## Respondent recruitment

We will use multi-channel respondent recruitment, including both phone and digital outreach. Phone outreach will allow us to capture older demographics and individuals who use the internet and social media less. Digital outreach, such as targeted social media ads, email lists, and news websites will enable us to recruit a broad sample. Digital outreach channels are cost effective and can reach younger demographics. We will allocate \$28,000 of our budget for rewarding survey respondents to improve our response rates, particularly within harder-to-reach groups such as young people and low-information voters [@tswd]. As shown by @BMC, incentives increase survey participation, particularly amongst lower-response demographics.

## Survey Design

Our survey is conducted using GoogleForms, which allows for an easy, user-friendly experience and increases respondent likelihood. The survey length is kept within the five to ten minute range, with thirteen questions to balance the need for data collection with the need for respondent engagement [@stantcheva2023].

The survey includes the following sections:

-   Demographics: In this section, we collect essential demographic data that we will use for post-survey weighting. This includes gender and ethnicity.

-   Candidate evaluation: This section includes direct questions about candidate support, vote choice, and support level.

-   Key issues and concerns: This section includes questions that aim to identify the top issues driving voter decisions, such as abortion rights and the economy.

## Data Validation

Common issues with surveys include incomplete or inconsistent responses, duplicate responses, and fake responses by bots. We will use automate validation checks to detect and filter out incomplete or inconsistent responses, IP tracking to prevent duplicate responses, and CAPTCHA technology to minimize bot interference [@zhang2019].

## Poll Aggregation

We will use the seven-day average method to aggregate polls because it decreases fluctuations in daily responses by smoothing out temporary spikes. This will make trends clearer.

## Weighting and Data Adjustments

We will use post-stratification weighting to correct demographic imbalances in our final sample [@nytsienacrosstabs].

## Budget

## Idealized survey questions

2024 U.S. Presidential Election Poll This survey collects information about voters' political views and who they support in the 2024 U.S. Presidential Election. The data collected will not be shared with any external parties and will strictly be used for analytical purposes. This survey is completely anonymous and your data will be protected. Any published material regarding the results drawn from this survey cannot be traced back to you. The goal of this survey is to draw conclusions about the 2024 presidential elections held in the United States. Please answer as accurately as possible. If you have any questions or concerns, please reach out to aliza.mithwani\@mail.utoronto.ca (correspondence will not be shared with any external parties).

1.  Are you a registered voter in the United States ? Yes No
2.  Do you plan to vote in the upcoming presidential election? Yes No Undecided

**Demographics Questions**

3.  Would you consider yourself: White Black or African American Hispanic or Latino Asian American Indian or Alaskan Native Middle Eastern or North African Native Hawaiian or Pacific Islander Prefer not to say Other (specify)

4.  What is your age? 18-29 20-44 45-64 65+ Prefer not to say

5.  What sex were you assigned at birth, on your original birth certificate? "Female" Male Prefer not to say

6.  How do you currently describe yourself (select all that apply)? "Female" Male Transgender Prefer not to say Other (Specify)

7.  What is your household income? Less than \$20,000 \$20,000-59,999 \$60,000-79,999 \$80,000-99,999 \$100,000 or more Prefer not to say

8.  In which state do you currently reside? Dropdown list of states

**Candidate Evaluation**

9.  If the 2024 presidential election were held today, who would you vote for?

Donald Trump, Republican Kamala Harris, Democrat Jill Stein, Green Party Write-in Don't know Prefer not to say Other (please specify)

(Optional) If you selected "Write-in" for the last question, please specify below:

10. Do you consider yourself a Democrat, a Republican, an Independent, or a member of another party? Democrat Republican Independent Another Party Don't know Prefer not to say

**Key Issues and Concerns**

11. Rate these issues in order of importance to you (1 being most important and 7 being the least important): (use multiple choice grid with 7 rows and 7 columns) Abortion Immigration The state of democracy/corruption Foreign policy The economy Character Climate change?
12. If you had to assign a value from 1 to 5 to your level of optimism about the future of the United States, where 1 means highly pessimistic and 5 means highly optimistic, where would you place yourself? 1 (Highly pessimistic) 2 3 4 5 (Highly optimistic)
13. What would you say to someone who is undecided about voting in this election?

**Confirmation Message**

Thank you for your response! Your answers have successfully been recorded.

**Link to survey:** <https://forms.gle/h7MTA8k21ZbYxahT6>

# Pollster methodology overview and evaluation {#sec-pollster-meth}

## Overview

The Siena College/New York Times (Times/Siena) poll is a collaboration between the New York Times and the Siena College Research Institute that aims to capture voter sentiment in battleground states and at a national level [@nytsiena]. Its accuracy has made it a highly regarded poll for predicting U.S. election results. The NYT/Siena poll uses live poll interviews to reach a representative sample of voters and measure their political preferences, views, and issue priorities [@nytsiena].

## Population and Sampling Frame

The target population, defined by @tswd is the population that the Times/Siena poll aims to speak about, is registered voters and likely voters in the United States. Some polls conducted by Times/Siena have a target population of registered voters, while others use likely voters [@nytsienacrosstabs]. The sampling frame, defined by @tswd as the individuals from the target population that Times/Siena can get data about is drawn from various voter registration databases, including the L2 voter database [@nytsiena]. In polls for the 2024 presidential election, Times/Siena has put an emphasis on gathering nationally-representative and state-specific samples for the seven battleground states that decided the 2020 presidential election and are expected to decide the 2024 presidential election: Arizona, Nevada, Georgia, North Carolina, Michigan, Wisconsin, and Pennsylvania [@nytsiena]. A sampling frame that includes various voter registration databases helps ensure precise targeting and that the sample reflects the demographics of registered or likely voters in the upcoming U.S. election [@nytsiena].

## Respondent Recruitment Strategy

Times/Siena recruits respondents for its poll at random, from a national list of registered voters gathered from various high-quality voter registration databases [@nytsiena].

## Sampling Approch and Trade-offs

## Strengths and Limitations of Methodology

### Strengths

### Limitations

# Additional data details

## Data Cleaning {#sec-data-cleaning}

## Polling averages from before President Joe Biden ended his re-election campaign {#sec-Biden}

<!-- It is impossible to quantify or model the effects of Biden's unprecedented decision to end his re-election campaign in July 2024. There is no established methodology for analyzing this, so we made the decision to focus on Harris versus Trump polling from after July 21, 2024 because they are the two major presidential nominees. For this reason, national and swing-state polling averages in the six months leading up to election day are shown here. -->

```{r}
#| label: fig-nationalavg
#| fig-cap: National popular vote averages for the Democratic and Republican presidential nominees since May 5, 2024 (six months before election day) show former President Trump leading before Biden ended his campaign, the gap narrowing after Harris became the Democratic nominee, Harris taking the lead in August, and a dead heat in September and October. 
#| echo: false
#| warning: false
#| message: false

national <- analysis_data |>
  filter(state=="National")

ggplot(national, aes(x = end_date, y = pct, color = party)) +
  stat_summary(geom = "line", fun = mean) +      # Plot the mean line for polling percentages
  geom_point(size = 1, alpha = 0.2) +            # Add points for each individual poll
  scale_color_manual(values = c("DEM" = "blue", "REP" = "red")) +
  geom_text(aes(x = as.Date("2024-07-21") + 1, y = 55,  # Adjust x to move the label to the right
                label = "Biden ends re-election campaign"), 
            color = "#A9A9A9", 
            vjust = -0.5,  # Adjust vertical position
            hjust = 0,     # Keep it left-aligned for better placement
            size = 2.5) + 
  geom_vline(xintercept = as.Date("2024-07-21"), linetype = "dashed", color = "#A9A9A9") +
  labs(title = "",
       x = "",
       y = "",
       color = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(limits = c(0, 60), labels = scales::percent_format(scale = 1)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %e")  # Use scale_x_date for date variables
```

```{r}
#| label: fig-sevenbattlegrounds
#| fig-cap: Democratic vs Republican presidential polling averages from May 5 to October 29, 2024, including both Biden and Harris as the Democratic nominee and Trump as the Republican nominee.
#| echo: false
#| warning: false
#| message: false
sevenbattlegrounds <- analysis_data |>
  filter(state=="Wisconsin" | state == "Michigan" | state == "Pennsylvania" |
           state == "Nevada" | state == "Arizona" | state == "Georgia" | 
           state == "North Carolina" | state == "Nebraska CD-2")

ggplot(sevenbattlegrounds, aes(x = end_date, y = pct, color = party)) +
  stat_summary(geom = "line", fun = mean) +      # Plot the mean line for polling percentages
  geom_point(size = 1, alpha = 0.15) +            # Add points for each individual poll
  scale_color_manual(values = c("DEM" = "blue", "REP" = "red")) +
  geom_text(aes(x = as.Date("2024-07-21") + 1, y = 55,  # Adjust x to move the label to the right
                label = "Biden ends re-election campaign"),
            color = "#A9A9A9",
            vjust = -0.5,  # Adjust vertical position
            hjust = 0,     # Keep it left-aligned for better placement
            size = 1.7) +
  geom_vline(xintercept = as.Date("2024-07-21"), linetype = "dashed", color = "#A9A9A9") +
  labs(title = "",
       x = "",
       y = "",
       color = "") +
  labs(title = "",
       x = "Date",
       y = "",
       color = "Party") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), size=1.5) +
  scale_y_continuous(limits = c(40, 60), labels = scales::percent_format(scale = 1)) +
  scale_x_date(date_breaks = "2 month", date_labels = "%b %e") +
  facet_wrap(~state)
```

# Model details {#sec-model-details}

Model summary is shown in @tbl-model-summary

```{r}
#| echo: false
#| eval: true
#| label: tbl-model-summary
#| tbl-cap: "Explanatory models of support for Harris and Trump based on end date (number of days since Biden ended re-election campaign on July 21, 2024), state, pollster, and pollscore"
#| warning: false

modelsummary::modelsummary(
  list(
    "Harris" = harris_model,
    "Trump" = trump_model
  ),
  statistic = "mad",
  fmt = 2,
  output = "kableExtra"    
) %>%
  kableExtra::kable_styling(font_size = 7) 
```

## Posterior predictive check

### Harris Model

In @fig-ppcheckandposteriorvspriorharris-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvspriorharris-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvspriorharris
#| layout-ncol: 2
#| fig-cap: "Examining how the model for support for Harris fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

# library(bayesplot)
# library(splines)
# library(ggplot2)
# 
# pp_check(harris_model) +
#   theme_classic() +
#   theme(legend.position = "bottom")
# 
# posterior_vs_prior(harris_model) +
#   theme_minimal() +
#   scale_color_brewer(palette = "Set1") +
#   theme(legend.position = "bottom") +
#   coord_flip()
```

### Trump Model

In @fig-ppcheckandposteriorvspriortrump-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvspriortrump-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvspriortrump
#| layout-ncol: 2
#| fig-cap: "Examining how the model for support for Trump, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

# library(bayesplot)
# library(splines)
# 
# pp_check(trump_model) +
#   theme_classic() +
#   theme(legend.position = "bottom")
# 
# posterior_vs_prior(trump_model) +
#   theme_minimal() +
#   scale_color_brewer(palette = "Set1") +
#   theme(legend.position = "bottom") +
#   coord_flip()
```

## Model Diagnostics

### Harris Model

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm for the Harris model"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 1

# Assuming harris_model is a Bayesian model, use bayesplot for flexible customization
library(bayesplot)
plot(harris_model, "trace")
plot(harris_model, "rhat")
```

### Trump Model

@fig-trumpdiagnostics-1 is a trace plot. It shows... This suggests...

@fig-trumpdiagnostics-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trumpdiagnostics
#| fig-cap: "Checking the convergence of the MCMC algorithm for the Harris model"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 1

plot(trump_model, "trace")
plot(trump_model, "rhat")
```

\newpage

# References
